
---
title: "climateStability Vignette: the Basics"
author:
- Hannah Owens
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{How to use climateStability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(climateStability);
library(raster);
library(rangeBuilder);
knitr::opts_chunk$set(error = TRUE);
```

# Introduction 

`climateStability` is a small set of `R`-based tools for generating climate stability estimates based on time slices of climate data.

# Calculating Stability
First, you give the `deviationThroughTime` function the location of a folder containing time slice data (as .ascii rasters) for a single variable, such as that produced by PaleoView (Fordham, *et al*. 2017, Ecography), and the time elapsed *between slices*. It will calculate the average standard deviation over time for each cell. Note that `deviationThroughTime` will use *all* the raster files in the given folder in order, so be careful that it only contains the files from which you want to calculate stability and that `R` loads the files in the correct order.

The function accomodates datasets with even temporal intervals, as well as uneven intervals. First, here is an example with a dataset that has time slices every 1,000 years.

```{r example of deviation through time with even time slices, eval=FALSE}
# deviationThroughTime using even time slices
precipDeviation <- deviationThroughTime(variableDirectory = "../../ClimateStabilityManuscript/precipfiles/",
                                        timeSlicePeriod = 1000);
temperatureDeviation <- deviationThroughTime(variableDirectory = "../../ClimateStabilityManuscript/tempfiles/",
                                             timeSlicePeriod = 1000);
```

Here is an example with a dataset that has uneven time slices for 1kybp to 5 kybp, 10kybp, 15kybp, and 21kybp. In this case, `timeSlicePeriod` can take a vector specifying the time elapsed between slices. NOTE: make sure the vector is in the *same order* as your time slices. There should be one fewer `timeSlicePeriod` item than the numer of climate slice files.

```{r example of deviation through time with uneven time slices, eval=FALSE}
# deviationThroughTime using uneven time slices
unevenPrecipDeviation <- deviationThroughTime(variableDirectory = "../../ClimateStabilityManuscript/precipfilesThinned/", 
                                              timeSlicePeriod = c(1000, 1000, 1000, 1000, 5000, 5000, 6000));
unevenTemperatureDeviation <- deviationThroughTime(variableDirectory = "../../ClimateStabilityManuscript/tempfilesThinned/",
                                                   timeSlicePeriod = c(1000, 1000, 1000, 1000, 5000, 5000, 6000));
```

You have been provided with two pre-calculated deviation-through-time estimates from the OWENS BIODIVERSITY INFORMATICS PAPER. These were calculated from 100-year climate means for 20 1,000-year time slices from 21 kbp to 1 kbp using data from the TRaCE 21k climate experiment, run in the CCSM 3.0 climate model. These layers will be used to illustrate the remaining `climateStability` package functions.

Let's load them into the `R` environment.

```{r load the results of deviation through time from the package}
#Using deviationThroughTime
data(precipDeviation, package = "climateStability");
data(temperatureDeviation, package = "climateStability");
```

The next step is scaling the deviation through time estimate from 0 to 1, and then take the inverse of the result. This how you estimate stability for a given climate variable, per OWENS BIODIVERSITY INFORMATICS PAPER.

```{r calculate stability from deviation}
precipInvDev <- 1/precipDeviation;
precipStability <- rescale0to1(precipInvDev);

tempInvDev <- 1/temperatureDeviation;
tempStability <- rescale0to1(tempInvDev);
```

Finally, to estimate climatic stability, multiply the individual climate variable stability estimates together.

```{r calculate climate stability from individual climate variable stability estimates}
climateStability <- rescale0to1(precipStability * tempStability)
```

# Plots showing stability

These data can be visualized either as rasters or as line graphs showing the relationship between latitude and stability (or however you want to display them--I won't tell you how to live your life).

## Stability rasters

This is what climate the stability rasters look like.

```{r plot stability rasters, fig.height=4, fig.width=8}
plot(precipStability, main = "Relative Precipitation Stability");
plot(gshhs, add = T)
plot(tempStability, main = "Relative Temperature Stability");
plot(gshhs, add = T)
plot(climateStability, main = "Overall Relative Climate Stability");
plot(gshhs, add = T)
```


## Stability line graphs

The `climateStability` package also comes with a useful tool for calculating mean stability per line of latitude, `latitudinalMean`. This can then be used to plot climate stability for a given variable as a response to latitude.

```{r plot latitudinal mean in stability, fig.height=4, fig.width=8}
#Calculate mean values at rasters
plot(latitudinalMean(precipStability), main = "Precipitation Stability by Latitude", 
     ylab = "Relative Stability", type = "l");

plot(latitudinalMean(tempStability), main = "Temperature Stability by Latitude", 
     ylab = "Relative Stability", type = "l");

plot(latitudinalMean(climateStability), main = "Climate Stability by Latitude", 
     ylab = "Relative Stability", type = "l");
```

Additionally, `climateStability` can calculate mean stability versus the absolute value of the latitudinal mean using the function `absLatitudinalMean`. 

```{r plot absolute latitudinal mean in stability, fig.height=4, fig.width=8}
#Calculate mean values at rasters
plot(absLatitudinalMean(precipStability), main = "Precipitation Stability by Absolute Latitude", 
     ylab = "Relative Stability", type = "l");

plot(absLatitudinalMean(tempStability), main = "Temperature Stability by Absolute Latitude", 
     ylab = "Relative Stability", type = "l");

plot(absLatitudinalMean(climateStability), main = "Climate Stability by Absolute Latitude", 
     ylab = "Relative Stability", type = "l");
```
